A standard for the scholarly citation of archaeological data as an incentive to data sharing

### 2017-11-11 analysis/paper/paper.Rmd
working on revisions

---
title: "A Standard for the Scholarly Citation of Archaeological Data"
author: 
  - name          : "Ben Marwick"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Department of Anthropology, University of Washington, Seattle, 98115, WA, USA"
    email         : "bmarwick@uw.edu"
  - name          : "Suzanne E. Pilaar Birch"
    affiliation   : "2"

@@

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
keywords          : ""
wordcount         : ""

@@

library(datacitation)

@@

We will start by defining what we mean by ‘data’, and reviewing previous work on researchers’ attitudes to sharing data. We motivate data sharing and citing behaviours by explaining why data sharing is beneficial to archaeology as a discipline, and to individual researchers. Then we will survey some of the data sharing and citing norms that have emerged in other research areas. From these norms we will present a standard suitable for citation of archaeological data. Finally, we will present some basic instructions on how to share archaeological data.

@@

This is a useful definition because it emphasizes data as digital manifestations, and so suitable for transmission, storage and retrieval using the internet. Physical objects (such as samples, collections, hard copies of notes, etc.) are data also, but they are out of scope of this discussion because of the different requirements for access, archiving, and so on. For many archaeologists, data are the contents of the files we use to generate visualisations, summary tables, and qualitative statements that appear in, and support the claims we make in, our scholarly publications and reports. The categories shown in Table \@ref(tab:types-of-data) are useful as a rough guide to what data can be, although this table is not comprehensive, and does not represent the full diversity of data used by archaeologists. Ultimately, data are whatever the community decides, as the The National Science Foundation notes in their commentary on Data Management Plans: "What constitutes such data will be determined by the community of interest through the process of peer review and program management. This may include, but is not limited to: data, publications, samples, physical collections, software and models." This is a highly pragmatic approach to defining data, but is also consistent with the relational account of data in philosophy of science scholarship, which holds that what data are is dependent on who uses it, how and for what purpose (Leonelli 2015; Daston and Galison 1992). It implies that the answer to the question of what counts as data is mostly emergent from our professional interactions, such as the materials we request when we do peer review of manuscripts and grant proposals, or when we make an informal request for more information by email.

Pragmatically in our own work, we consider that the main data suitable for sharing as the contents of the computer files (e.g. text, raw structured data, and images) that we analyse or otherwise use to create the figures, tables and quantitative results that we include in our reports and publications. The selection omits materials that could also be considered data according to the definitions noted above (e.g. field notes), but we feel our selection represents the most information-dense and meaningful data records that are useful for others wanting to assess the reliability of our work, or reuse our findings in their own work. These are also the items that have been most frequently requested by others writing to us about our publications. Our citation standard imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion. Table \@ref(tab:desirable) summarises some desirable minimum qualities for archaeological data that optimise its discovery and reuse.

```{r types-of-data}
tbl <- readr::read_csv("../figures/tbl_types_of_data_nsb.csv")
knitr::kable(tbl, caption = "A selection of types of data encountereed by archaeologists.")
```

```{r desirable}
tbl <- readr::read_csv("../figures/tbl_rules_of_data_sharing.csv")
knitr::kable(tbl, caption = "Best practices for preparing data for sharing, from Hart et al. (2016) ")
```

@@

While public and community benefits and costs are well-rehearsed and widely agreed on (Table \@ref(tab:costsbenefits)), they offer no, or only weak, tangible incentives to the individual researcher (Rowhani-Farid et al. 2017). Previous work has identified citation advantages as a tangible benefit to individual researchers who share data with publications, although the magnitude of the advantage varies greatly (see Table \@ref(tab:indivbenefits) for a summary). Increased publication productivity also appeared to be associated with data sharing behaviours. Across over 7,000 NSF and NIH awards, Pienta et al. (2010) reported that research projects with archived data produced a median of 10 publications, versus only five for projects without archived data.

```{r costsbenefits}
tbl <- readr::read_csv("../figures/tbl_reasons_to_share_data.csv")
knitr::kable(tbl, caption = "Summary of public and community benefits and costs of data sharing")
```

@@

Stodden (2010) surveyed registrants at a top machine learning conference (the Neural Information Processing Systems (NIPS) conference) to identify the factors that affect researchers’ decisions to reveal data (Table \@ref(tab:costsbenefits)). Stodden found that communitarian factors influence scientists to share their data, but private incentives determine when scientists choose not to share their code and data. In the top three reasons not to share, we see a lack of incentives to invest time in data sharing (however, there are many other time-consuming tasks in research, and they get done because clear reward structures are in place); a lack of faith in data citation (which we hope this paper will help with); and a lack of an efficient copyright process. Stodden’s (2009a, b) ‘Reproducible Research Standard’, summarised in Figure \@ref(fig:stodden-rrs), is a convenient and robust copyright schema that is in wide use for many typical research projects.

```{r stodden-rrs, fig.cap="Summary of Stodden’s (2009b) 'Reproducible Research Standard'"}
knitr::include_graphics("../figures/fig1_rrs.png")
```

@@

Our sample focused on _Journal of Archaeological Science_ articles on experiments with stone artefacts that were published during `r min(email_survey$'publication year', na.rm = TRUE)` and `r max(email_survey$'publication year', na.rm = TRUE)`. We focused on stone artefact experiments because all research materials in these projects are generated by the researcher, so there are no other stakeholders to consider when determining how to share the data: the researcher has complete control over the data. There are also no risks of damage to people or property in sharing these data because they do not include archaeological sites or human subjects. In contacting the authors, we requested the datasets for use in a graduate seminar with the aim of reproducing the published results and further exploring the data (this stage is still in progress). We promised to contact the authors if we discovered anything new, and not to share the data further without their permission. The data included with this paper is anonymised so the specific authors we contacted are not identifiable.

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and received `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We received `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We interpret these to mean that the author does not have time to prepare their data for sharing, perhaps because the data are not already in a well-organised form ready for distribution. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing.

@@

We see similar rates of sharing in response to email requests among archaeologists to what have been described in other disciplines (Table \@ref(tab:sharing-rates), Tenopir et al. 2015; Vanpaemel et al. 2015). However, our result may overestimate the true rate because our sample did not include datasets that authors might not be able or willing to share because they contain sensitive location or cultural information. Analysis of the qualitative responses suggests two key reasons why authors are reluctant to share: they are fearful that they will lose the opportunity to produce further publications from those data (a typical concern of researchers in data-poor fields), and that their data are not in a form suitable for sharing and the task of organising those data is not a high priority. These reasons not to share have also been documented in other research areas, noted above. The absence of patterns in the date of the paper, student status of the author and other variables may relate to more to our small sample size than a real absence of relationships in data sharing behaviour.

@@

```{r date-and-sharing, fig.cap="There is no obvious trend in data sharing becavious and date of publication"}
library(viridis)
years <- seq(min(email_survey$'publication year', na.rm = TRUE),
    max(email_survey$'publication year', na.rm = TRUE), 1)

sharing_by_year <- 
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
  group_by(`publication year`, `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`publication year`)) %>% 
  mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  ggplot(aes(`publication year`, n)) +
    geom_col(aes(fill = `content recieved date`),
             # need devtools::install_github("tidyverse/ggplot2") for this
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = FALSE) +
    scale_x_continuous(breaks = years,
                       labels = years) +
  xlab("Publication year") +
    theme_bw() +
  coord_fixed(ratio = 0.5)

@@

```{r student, fig.cap="No relationship between student status and data sharing behaviour"}
sharing_by_student <- 
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  `author is graduate student at time of publication?`, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`author is graduate student at time of publication?`)) %>% 
  ggplot(aes(`author is graduate student at time of publication?`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = FALSE) +
    xlab("Author was graduate student at time of publication?") +
    theme_bw() +
  coord_fixed(ratio = 0.05)
```

```{r author-country}
sharing_by_country <- 
email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  `author home`, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(`author home`)) %>% 
  ggplot(aes(`author home`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = FALSE) +
    theme_bw() +
  xlab("Author location at time of publication") +
  coord_fixed(ratio = 0.25)
```

```{r journals, fig.cap="No relationship between sharing behaviour and journal."}
sharing_by_journal <- 
  email_survey %>% 
  filter(!is.na( `email sent date`)) %>% 
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`), 
  "yes", "no")) %>% 
  group_by(
  journal, 
  `content recieved date`) %>% 
  tally() %>% 
  filter(!is.na(journal)) %>% 
  ggplot(aes(journal, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
    scale_fill_viridis(discrete = TRUE,
                       guide = guide_legend(title = "Data shared?")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = c(0.8, 0.7)) +
  xlab("Journal name") +
  coord_fixed(ratio = 0.25)
```

@@

```{r fig.cap="Summary of results of email requests for data. A. Responses by year of journal article publication, B. Student status of first author, C. Country location of first author, D. Journal (AAS = Archaeological and Anthropological Sciences, AMANT = American Antiquity, ARCHMTRY = Archaeometry, JAS = Journal of Archaeological Science, JFA = Journal of Field Archaeology, JHE = Journal of Human Evolution, LT = Lithic Technology, SCI = Science."}
knitr::include_graphics("../figures/email_sharing_plots_hand_edit.png")
```

@@

In addition to our email survey, we also conducted a pilot study of a random sample of `r nrow(jas_article_survey)` articles published during 2016-2017 in the _Journal of Archaeological Science_ to investigate data sharing behaviours. In this sample we found openly available raw data in `r data_are_avail` papers (`r data_are_avail_perc`%), even though only `r have_data_avail_statement` articles (`r have_data_avail_statement_perc`%) include a data availability statement. For papers where data was available, `r data_in_supp` papers (`r data_in_supp_perc`%) have data contained in supplementary files published with the paper, `r data_in_table` (`r data_in_table_perc`%) have all the raw data in tables in the text of the paper, and `r data_in_repo` (`r data_in_repo_perc`%) have data in an online repository. The most frequently shared type of data is compositional (e.g. element concentrations, n = `r comp_n`, `r comp_perc`%), followed by radiocarbon and luminescence age data (where the dates are the primary object of analysis, n = `r dates_n`, `r dates_perc`%) and DNA sequences (n = `r dna_n`, `r dna_perc`%). Generally, compositional data is presented in tables in the text, dating data is presented in supplementary files, and DNA data are in repositories (as is typical for DNA data). Data shared in supplementary files and online repositories were most often in Microsoft Excel format (n = `r xlsx_n`, `r xlsx_perc`% of the files in supplements files and repositories), followed by CSV (comma separated variables, an open source plain text spreadsheet format, n = `r csv_n`, `r csv_perc`%), and tables embedded in Microsoft Word or PDF documents (n = `r doc_n`, `r doc_perc`%). When a scripting language such as R, Python, MATLAB or OxCal was used as the analysis software for a paper (n = `r script_n`, `r script_perc`% of the papers where would identify the software), only 2 papers also provided the script with the paper.

The _Journal of Archaeological Science_ has had a 'data disclosure' policy since at least 2013 (Editors, 2013) which states that “All data relating to the article must be made available in Supplementary files or deposited in external repositories and linked to within the article”. The data availability rate of 53% reported in our pilot study reflects either weak enforcement of this policy or an incomplete understanding among editors and authors of how to interpret and implement this policy. The prominence of compositional and dating data among the shared data types suggests a least effort strategy by authors sharing data that do not require extensive cleaning and tidying (because the data are generated by instruments rather than entered by hand, such as artefact attribute data). The size of compositional and dating datasets means that they mostly fit easily into tables in the text or supplementary files, and do not require much effort to organise (e.g. very little metadata is required for others to use these data). The small proportion of papers in our sample using or sharing a scripting language suggests that unlike other fields (Eglen et al. 2017), archaeologists do not appear to recognise code for data analysis as a primary research object to share with the research community.

@@

plot_file_type_location <- 
file_type_location %>% 
  group_by(file, location) %>% 
  tally() %>% 
  ggplot(aes(file, location)) + 
  geom_point(aes(size = n),
             colour = viridis(2)[2],
             shape = 16) +
  theme_bw() + 
  xlab("") + 
  ylab("") +
  scale_size_continuous(range=c(7,25)) + 
  geom_text(aes(label = n)) +
  theme(legend.position = "none") 
```

@@

plot_data_type_location <- 
data_type_location %>% 
  group_by(data_type, location) %>% 
  tally() %>% 
  filter(n >= 2) %>% 
  ggplot(aes(data_type, location)) + 
  geom_point(aes(size = n),
             colour = viridis(2)[1],
             shape = 16) +
  theme_bw() + 
  xlab("") + 
  ylab("") +
  scale_size_continuous(range=c(7,25)) + 
  geom_text(aes(label = n), colour = "white") +
  theme(legend.position = "none") 
```

```{r  fig.cap="Summary of relationships between file types, data types, and data locations. A. File types and data locations, B. Data types and data locations. "}
plot_data_type <- plot_grid(plot_file_type_location, 
                            plot_data_type_location,
                      labels="AUTO", ncol = 1)
plot_data_type
```

@@

library(viridis)
date_plot <- 
  dates  %>% 
  filter(!is.na(yyyy_date)) %>% 
  ggplot(aes(yyyy_date,
             fill = datacentre
             )) +
  geom_bar() +
  # guides(fill=FALSE) +
  scale_fill_viridis(discrete = TRUE,
                     breaks = top_data_sources[1:5],
                     labels = top_data_sources[1:5]) +
  theme_new() +
  theme(legend.position = c(0.3, 0.5)) +
  scale_x_continuous(limits = c(1950,2017), expand = c(0,0)) +
  xlab("Year data was collected") +
  ylab("Number of archaeology DOIs\nknown to DataCite") +
  geom_vline(xintercept = peak_year, colour = "red") +
  annotate("text", 
           x = peak_year - 1.5,
           y = ymax,
           label = peak_year,
           size = 3) # +
  # coord_fixed(ratio = 1/1000)

# change in each major repo over time
per_repo <- 
dates  %>% 
  filter(!is.na(yyyy_date)) %>% 
  group_by(yyyy_date, datacentre)  %>% 
  tally() %>% 
  filter(n != 0,
         datacentre %in% top_data_sources[1:9]) %>% 
    mutate(datacentre = str_wrap(datacentre, width = )) %>% 
   ggplot(aes(yyyy_date,
              n
             )) +
  geom_line(size = 1)  +
  scale_color_viridis(discrete = TRUE) +
  scale_y_log10() +
  xlim(c(1990, 2017)) +
  xlab("Trends over time at major archaeological data repositories") +
  theme_new() +
  facet_wrap(~ datacentre)

@@

```{r eval = FALSE}
# text mine a little
library(tidytext)
subject_1gram <- 
xx %>% 
  unnest_tokens(word, subject) %>%
  anti_join(stop_words) %>%
  count(word, sort = TRUE)

subject_2gram <- 
xx %>% 
  unnest_tokens(bigram, subject, token = "ngrams", n = 2) %>%
  count(word, sort = TRUE)

subject_3gram <- 
xx %>% 
  unnest_tokens(trigram, subject, token = "ngrams", n = 3) %>%
  count(word, sort = TRUE) 
```

>>>>

---
title: "A standard for the scholarly citation of archaeological data as an incentive to data sharing"
author:
  - name          : "Ben Marwick"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Department of Anthropology, University of Washington, Seattle, 98115, WA, USA"
    email         : "bmarwick@uw.edu"
  - name          : "Suzanne E. Pilaar Birch"
    affiliation   : "2"

@@

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
keywords          : ""
wordcount         : '`r  unname(stringi::stri_stats_latex(readLines("paper.Rmd"))[4])` (includes reference list )'

@@

library(datacitation)
library(tidyverse)

@@

Archaeologists work with data in research, commercial, government, and other contexts. A full investigation of data in each of these contexts is beyond the scope of this paper, so we limit this study to what we are most famliar with, the research context. However, many of the practical details we discuss here will also be relevant to archaeologists working in other contexts. We begin by defining what we mean by ‘data’, and reviewing previous work on researchers’ attitudes to sharing data.  We motivate data sharing and citing behaviours by explaining why data sharing is beneficial to archaeology as a discipline, and to individual researchers. Then we survey some of the data sharing and citing norms that have emerged in other research areas. From these norms we derive a standard suitable for citation of archaeological data. Finally, we present some basic instructions on how to share archaeological data.

@@

This is a useful definition because it emphasizes data as digital manifestations, and so suitable for transmission, storage and retrieval using the internet. Physical objects (such as samples, collections, hard copies of notes, etc.) are data also, but they are out of scope of this discussion because of the different challenges and requirements for access, archiving, and so on. The categories shown in Table \@ref(tab:types-of-data) are useful as a rough guide to what data can be, although this table is not comprehensive, and does not represent the full diversity of data used by archaeologists. Ultimately, data are whatever the community decides, as the The National Science Foundation notes in their commentary on Data Management Plans: "What constitutes such data will be determined by the community of interest through the process of peer review and program management. This may include, but is not limited to: data, publications, samples, physical collections, software and models." These definitions in policy are consistent with the relational account of data in philosophy of science scholarship, which holds that what data are is dependent on who uses it, how and for what purpose (Leonelli 2015; Daston and Galison 1992). How can these definitions and philosophical discussions inform our practice? They tell us that what counts as data is emergent from our professional interactions. Data are the materials we request when we do peer review of manuscripts and grant proposals, or when we make an informal request for more information by email. Ther are not exclusively objective archaeometric records, but include interpretive and derivative results. In short, data are anything another person needs to understand our work better.

To translate this into a more applied context, our practice of data sharing it to make available the computer files (e.g. containing text, raw structured data, images, etc.) that we analyse or otherwise use to create the figures, tables and quantitative results that we include in our reports and publications. These include our derived results and secondary data, such as taxonomic and typological classifications. This selection omits materials that could also be considered data according to the definitions noted above (e.g. field notes), but we feel our selection represents the most information-dense and meaningful items that are useful for our community of interest. These are also the items that have been most frequently requested from us by others in our professional interactions. Our citation standard is inclusive and imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion.

```{r types-of-data}
tbl <- readr::read_csv("../figures/tbl_types_of_data_nsb.csv")
knitr::kable(tbl, caption = "A selection of types of data encountered by archaeologists.")
```

@@

While public and community benefits and costs are well-rehearsed and widely agreed on (Tables \@ref(tab:sharing-costs) and \@ref(tab:sharing-benefits)), they offer no, or only weak, tangible incentives to the individual researcher (Rowhani-Farid et al. 2017). Previous work has identified citation advantages as a tangible benefit to individual researchers who share data with publications, although the magnitude of the advantage varies greatly (see Table \@ref(tab:indivbenefits) for a summary). Increased publication productivity also appeared to be associated with data sharing behaviours. Across over 7,000 NSF and NIH awards, Pienta et al. (2010) reported that research projects with archived data produced a median of 10 publications, versus only five for projects without archived data.

```{r sharing-costs}
tbl <- readr::read_csv("../figures/tbl_reasons_not_to_share_data.csv")
knitr::kable(tbl, caption = "Summary of public and community costs of data sharing, from Stodden (2010) and Borgman (2012)")
```

```{r sharing-benefits}
tbl <- readr::read_csv("../figures/tbl_reasons_to_share_data.csv")
knitr::kable(tbl, caption = "Summary of public and community benefits of data sharing, from Stodden (2010) and Borgman (2012)")
```

@@

Stodden (2010) surveyed registrants at a top machine learning conference (the Neural Information Processing Systems (NIPS) conference) to identify the factors that affect researchers’ decisions to reveal data (Table \@ref(tab:costsbenefits)). Stodden found that communitarian factors influence scientists to share their data, but private incentives determine when scientists choose not to share their code and data. In the top three reasons not to share, we see a lack of incentives to invest time in data sharing (however, there are many other time-consuming tasks in research, and they get done because clear reward structures are in place); a lack of faith in data citation (which we hope this paper will help with); and a lack of an efficient copyright process. Stodden’s (2009a, b) ‘Reproducible Research Standard’, summarised in Table \@ref(tab:stodden-rrs), is a convenient and robust copyright schema that is in wide use for many typical research projects.

```{r stodden-rrs, fig.cap="Summary of Stodden’s (2009b) 'Reproducible Research Standard'"}

tbl <- readr::read_csv("../figures/tbl_rrs.csv")
names(tbl) <- c("Contents", "Example", "License", "Comment")
knitr::kable(tbl, caption = "Summary of Stodden’s (2009b) 'Reproducible Research Standard'. This standard realigns legal rights with scientific norms by freeing scientific work from copying and reuse restrictions, with attribution. It encourages researchers to release their products with standardised instructions for reuse and attribution")
```

@@

Our sample focused on _Journal of Archaeological Science_ articles on experiments with stone artefacts that were published during `r min(email_survey$'publication year', na.rm = TRUE)` and `r max(email_survey$'publication year', na.rm = TRUE)`. We focused on stone artefact experiments because all research materials in these projects are generated by the researcher, so there are no other stakeholders to consider when determining how to share the data: the researcher has complete control over the data. There are also no risks of damage to people or property in sharing these data because they do not include archaeological sites or human subjects. Because we contacted authors of papers at the intersection of (1) containing eminently sharable data and (2) our _bona fide_ research interests (i.e. stone artefacts for BM), this is a targeted, non-random sample that may not be representative of the entire archaeological community. In contacting the authors, we requested the datasets for use in a graduate seminar with the aim of reproducing the published results and further exploring the data (this stage is still in progress). We promised to contact the authors if we discovered anything new, and not to share the data further without their permission. The data included with this paper is anonymised so the specific authors we contacted are not identifiable.

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and received `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We received `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing.

@@

We see similar rates of sharing in response to email requests among archaeologists to what have been described in other disciplines (Table \@ref(tab:sharing-rates), Tenopir et al. 2015; Vanpaemel et al. 2015). However, our result may overestimate the true rate because our sample did not include datasets that authors might not be able or willing to share because they contain sensitive location or cultural information. Analysis of the qualitative responses we received suggests two key reasons why authors are reluctant to share: they are fearful that they will lose the opportunity to produce further publications from those data (a typical concern of researchers in data-poor fields), and that their data are not in a form suitable for sharing and the task of organising those data is not a high priority. These reasons not to share have also been documented in other research areas, noted above. The absence of patterns in the date of the paper, student status of the author and other variables (Figure \@ref(fig:email-figs)) may relate to more to our small sample size than a real absence of relationships in data sharing behaviour.

@@

```{r date-and-sharing, fig.cap="There is no obvious trend in data sharing becavious and date of publication"}
library(viridis)
library(scales) 
library(wesanderson)

years <- seq(min(email_survey$'publication year', na.rm = TRUE),
    max(email_survey$'publication year', na.rm = TRUE), 1)

sharing_by_year <-
email_survey %>%
  filter(!is.na(`email sent date`)) %>%
  group_by(`publication year`, `content recieved date`) %>%
  tally() %>%
  filter(!is.na(`publication year`)) %>%
  mutate(`content recieved date` = if_else(!is.na(`content recieved date`),
  "yes", "no")) %>%
  ggplot(aes(`publication year`, n)) +
    geom_col(aes(fill = `content recieved date`),
             # need devtools::install_github("tidyverse/ggplot2") for this
             position = position_dodge(preserve = 'single')) +
  scale_fill_manual(values = wes_palette("Rushmore")[c(1,3)], 
                    guide = FALSE) +
  scale_x_continuous(breaks = years,
                    labels = years) +
  xlab("Publication year") +
  scale_y_continuous("Number of papers",
                     breaks= pretty_breaks()) +
  theme_bw() +
  coord_fixed(ratio = 0.5)

@@

```{r student, fig.cap="No relationship between student status and data sharing behaviour"}
sharing_by_student <-
email_survey %>%
  filter(!is.na( `email sent date`)) %>%
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`),
  "yes", "no")) %>%
  group_by(
  `author is graduate student at time of publication?`,
  `content recieved date`) %>%
  tally() %>%
  filter(!is.na(`author is graduate student at time of publication?`)) %>%
  ggplot(aes(`author is graduate student at time of publication?`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
  scale_fill_manual(values = wes_palette("Rushmore")[c(1,3)],
                    guide = FALSE) +
  scale_y_continuous("Number of papers",
                     breaks= pretty_breaks()) +
    xlab("Author was graduate student at time of publication?") +
    theme_bw() +
  coord_fixed(ratio = 0.05)
```

```{r author-country}
sharing_by_country <-
email_survey %>%
  filter(!is.na( `email sent date`)) %>%
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`),
  "yes", "no")) %>%
  group_by(
  `author home`,
  `content recieved date`) %>%
  tally() %>%
  filter(!is.na(`author home`)) %>%
  ggplot(aes(`author home`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
  scale_fill_manual(values = wes_palette("Rushmore")[c(1,3)],
                    guide = FALSE) +
    scale_y_continuous("Number of papers",
                     breaks= pretty_breaks()) +
    theme_bw() +
  xlab("Author location at time of publication") +
  coord_fixed(ratio = 0.25)
```

```{r journals, fig.cap="No relationship between sharing behaviour and journal."}

email_survey <- 
  email_survey %>% 
  mutate(journal = ifelse(journal == "AMANT", 
                          "AQ",
                          journal))

sharing_by_journal <-
  email_survey %>%
  filter(!is.na( `email sent date`)) %>%
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`),
  "yes", "no")) %>%
  group_by(
  journal,
  `content recieved date`) %>%
  tally() %>%
  filter(!is.na(journal)) %>%
  ggplot(aes(journal, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
  scale_fill_manual(values = wes_palette("Rushmore")[c(1,3)],
                    guide = guide_legend(title = "Data shared?")) +
    scale_y_continuous("Number of papers",
                     breaks= pretty_breaks()) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = c(0.8, 0.7)) +
  xlab("Journal name") +
  coord_fixed(ratio = 0.25)
```

@@

```{r email-figs, fig.cap="Summary of results of email requests for data. A. Responses by year of journal article publication, B. Student status of first author, C. Country location of first author, D. Journal (AAS = Archaeological and Anthropological Sciences, AQ = American Antiquity, ARCHMTRY = Archaeometry, JAS = Journal of Archaeological Science, JFA = Journal of Field Archaeology, JHE = Journal of Human Evolution, LT = Lithic Technology, SCI = Science."}
knitr::include_graphics("../figures/email_sharing_plots_hand_edit.png")
```

@@

```{r}
# what is the total number of articles published in the population that we drew our sample from

# start http://doi.org.offcampus.lib.washington.edu/10.1016/j.jas.2016.11.003
# end   http://doi.org.offcampus.lib.washington.edu/10.1016/j.jas.2017.04.002

# start: Volume 78, February 2017, Pages 66-77
# end:   Volume 81, May 2017, Pages 90-100

# Vol, n_papers
# 81, 11
# 80,  9
# 79, 10
# 78, 18

# total: 48

In addition to our email survey, we also conducted a pilot study of a random sample of `r nrow(jas_article_survey)` 48 articles published during February-May 2017 in the _Journal of Archaeological Science_ to investigate data sharing behaviours. In this sample we found openly available raw data in `r data_are_avail` papers (`r data_are_avail_perc`%), even though only `r have_data_avail_statement` articles (`r have_data_avail_statement_perc`%) include a data availability statement. For papers where data was available, `r data_in_supp` papers (`r data_in_supp_perc`%) have data contained in supplementary files published with the paper, `r data_in_table` (`r data_in_table_perc`%) have all the raw data in tables in the text of the paper, and `r data_in_repo` (`r data_in_repo_perc`%) have data in an online repository. The most frequently shared type of data is compositional (e.g. element concentrations, n = `r comp_n`, `r comp_perc`%), followed by radiocarbon and luminescence age data (where the dates are the primary object of analysis, n = `r dates_n`, `r dates_perc`%) and DNA sequences (n = `r dna_n`, `r dna_perc`%). Generally, compositional data is presented in tables in the text, dating data is presented in supplementary files, and DNA data are in repositories (as is typical for DNA data) (Figure \@ref(fig:data-types-and-files)). Data shared in supplementary files and online repositories were most often in Microsoft Excel format (n = `r xlsx_n`, `r xlsx_perc`% of the files in supplements files and repositories), followed by CSV (comma separated variables, an open source plain text spreadsheet format, n = `r csv_n`, `r csv_perc`%), and tables embedded in Microsoft Word or PDF documents (n = `r doc_n`, `r doc_perc`%). When a scripting language such as R, Python, MATLAB or OxCal was used as the analysis software for a paper (n = `r script_n`, `r script_perc`% of the papers where would identify the software), only 2 papers also provided the script with the paper.

The _Journal of Archaeological Science_ has had a 'data disclosure' policy since at least 2013 (Editors, 2013) which states that "All data relating to the article must be made available in Supplementary files or deposited in external repositories and linked to within the article". The data availability rate of `r data_are_avail_perc`% reported in our pilot study reflects either weak enforcement of this policy or an incomplete understanding among editors, reviewers, and authors of how to interpret and implement this policy. The prominence of compositional and dating data among the shared data types suggests a least effort strategy by authors sharing data that do not require extensive cleaning and tidying (because the data are generated by instruments rather than entered by hand, such as artefact attribute data). The size of compositional and dating datasets means that they mostly fit easily into tables in the text or supplementary files, and do not require much effort to organise (e.g. very little metadata is required for others to use these data). The small proportion of papers in our sample using or sharing a scripting language suggests that unlike other fields (Eglen et al. 2017), archaeologists do not appear to recognise code for data analysis as a primary research object to share with the research community. Code for even the simplest operations should be shared along with data because code communicates decisions made about the data analysis that can help others understand, analyse, and reuse the data.

@@

plot_file_type_location <-
file_type_location %>%
  group_by(file, location) %>%
  tally() %>%
  ggplot(aes(file, location)) +
  geom_point(aes(size = n),
             colour = wes_palette("Rushmore")[c(2)],
             shape = 16) +
  theme_bw() +
  xlab("") +
  ylab("") +
  scale_size_continuous(range=c(7,25)) +
  geom_text(aes(label = n)) +
  theme(legend.position = "none")
```

@@

plot_data_type_location <-
data_type_location %>%
  group_by(data_type, location) %>%
  tally() %>%
  filter(n >= 2) %>%
  ggplot(aes(data_type, location)) +
  geom_point(aes(size = n),
             colour = wes_palette("Rushmore")[c(2)],
             shape = 16) +
  theme_bw() +
  xlab("") +
  ylab("") +
  scale_size_continuous(range=c(7,25)) +
  geom_text(aes(label = n), colour = "black") +
  theme(legend.position = "none")
```

```{r data-types-and-files, fig.cap="Summary of relationships between file types, data types, and data locations. A. File types and data locations, B. Data types and data locations. "}
plot_data_type <- plot_grid(plot_file_type_location,
                            plot_data_type_location,
                      labels="AUTO", ncol = 1)
plot_data_type
```

@@

date_plot <-
  dates  %>%
  filter(!is.na(yyyy_date)) %>%
  filter(datacentre %in% top_data_sources[1:5]) %>% 
  ggplot(aes(yyyy_date,
             fill = datacentre
             )) +
  geom_bar() +
  # guides(fill=FALSE) +
  scale_fill_manual(values = wes_palette("Rushmore")) +
  theme_new() +
  theme(legend.position = c(0.3, 0.5)) +
  scale_x_continuous(limits = c(1950,2017), expand = c(0,0)) +
  xlab("Year data was collected") +
  ylab("Number of archaeology DOIs\nknown to DataCite") +
  geom_vline(xintercept = peak_year, colour = "red") +
  annotate("text",
           x = peak_year - 1.5,
           y = ymax,
           label = peak_year,
           size = 3) # +
  # coord_fixed(ratio = 1/1000)

# change in each major repo over time
per_repo <-
dates  %>%
  filter(!is.na(yyyy_date)) %>%
  group_by(yyyy_date, datacentre)  %>%
  tally() %>%
  filter(n != 0,
         datacentre %in% top_data_sources[1:9]) %>%
    mutate(datacentre = str_wrap(datacentre, width = )) %>%
   ggplot(aes(yyyy_date,
              n
             )) +
  geom_line(size = 1)  +
  scale_fill_manual(values = wes_palette("Rushmore")) +
  scale_y_log10() +
  xlim(c(1990, 2017)) +
  xlab("Year DOI registered (1990-2017 only)") +
  ylab("Number of DOIs\nregistered per year") +
  theme_new() +
  facet_wrap(~ datacentre)

@@
